\chapter{Conclusions}
\section{Have the assumptions been met?}
    In conclusion, the aim of this thesis has been achieved by creating a program -- \textit{MARS App} -- that fulfills this work's assumptions. 83\% of the code behind the app is an F\# code and, wherever possible, functional approach was used. Therefore, the functional part is covered without a doubt.
    
    Subsequently, derivative financial instrument in the form of a European-style option has been implemented and priced under the \textit{Black-Scholes Model} as presented in fig. \ref{fig:optionsPresentation}. Underlying asset's history has been generated using \textit{Geometric Brownian Motion}, as presented in the previous chapter. Ability to specialize the product has been introduced as well, what can bee seen in the fig. \ref{fig:parametersPresentation}. Elements of the graphic design concerning product specialization, option presentation and \textit{GBM} visualisation can be seen in corresponding sections.
    
    This way all assumptions have been met and several conclusions can be drawn from this work:
    \begin{itemize}
        \item Complex mathematical models can be introduced in few lines of F\# code.
        \item For the time being (December 2020) F\# language is not the best choice for creating a desktop application due to not large enough community, lack of tutorials and overall onerousness related to binding UI with back-end logic.
        \item \textit{Black--Scholes Model}, although nearly 50-year-old, is still widely used despite the era of Artificial Intelligence and its benefits.
    \end{itemize}
    
\section{Further development possibilities}
    Future works may concern adding:
    \begin{itemize}

        \item connection to an online database in order to use historical data for an underlying asset instead of artificial trends generated by \textit{GBM},
        \item so-called \textit{greeks} to further improve \textit{MARS App} utility,
        \item \textit{Monte Carlo Simulation} (written in C++ to speed up computing) as another model for comparison with \textit{GBM}.
    \end{itemize}